---
title: "Untitled"
author: "Nurit Edri"
date: "1/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

######## Adapted from Depmap vignette and "Exploiting Depmap cancer dependency data using the depmap R package" by Theo Killian and Laurent Gatto

Depmap, A "cancer dependency map", represents the "landscape" of *in-vitro* characterization of cancer cell lines (~1700). It provides exploring data of cancer dependencies from large-scale RNAi and CRISPR-Cas9 screens. These RNAi and CRISPR-Cas9 experiments aimed to identify genes essential for cancer cell proliferation/survival. The gene's essentiality across the different cell lines is represented as a dependency score. A lower score that a cell line is highly dependent (essential) on that gene.


*depmap* package in R imports Depmap data into:

* “rnai” (RNAi genetic dependency)

* “crispr” (CRISPR genetic dependency)

* “copyNumber” (log fold copy number)

* “TPM” (protein-coding expression)

* “RPPA” (Reverse Phase Protein Array)

* “mutationCalls” (mutation calls)

* “drug_sensitivity” (chemical dependency)

* “metadata” (metadata about all cancer cell lines)

```{r}
# Load packages
# Please install missing packages if needed

library("dplyr")
library("ggplot2")
library("viridis")
library("tibble")
library("gridExtra")
library("stringr")
library("depmap")
library("ExperimentHub")
```

Depmap data stored in the cloud via ExperimentHub. ExperimentHub is a Bioconductor's service which stores records from experiments, publications or training courses. 

We can access Depmap data using *ExperimentHub* package or directly by using *depmap* package.

```{r}
# An example for accessing Depmap data using *ExperimentHub* package 

eh <- ExperimentHub()
query(eh, "depmap")

rnai <- eh[["EH3080"]]
crispr <- eh[["EH3081"]]
mutationCalls <- eh[["EH3085"]]
metadata <- eh[["EH3086"]]
TPM <- eh[["EH3084"]]
copyNumber <- eh[["EH3082"]]
drug_sensitivity_eh <- eh[["EH3087"]]
```

### Mining, manipulation and visualization

For data manipulation we will use *dplyr* package:

*`mutate()` adds new variables that are functions of existing variables

*`select()` picks variables based on their names.

*`filter()` picks cases based on their values.

*`summarise()` reduces multiple values down to a single summary.

*`arrange()` changes the ordering of the rows.



### Home work
1. Make a scatter plot of expression vs rnai gene dependency for Rhabdomyosarcoma Sarcoma.

```{r}

TPM_rnai <- metadata  %>%
              dplyr::select(depmap_id, lineage) %>%
              dplyr::full_join(rnai , by = "depmap_id") %>%
  dplyr::filter(lineage == "rhabdomyosarcoma" , gene_name == "RPL14")  %>%
  dplyr::full_join(TPM %>% dplyr::select(depmap_id, expression), by = "depmap_id") %>%
  dplyr::filter(lineage == "rhabdomyosarcoma" , gene_name == "RPL14")  
              
p2 <- TPM_rnai %>%
      ggplot(aes(x = expression, y = dependency)) +
      geom_point(alpha = 0.4, size = 0.5)
p2
             
```

2. Make a boxplot displaying log genomic copy number for gene RPL14 by lineage

```{r}

metadata %>% dplyr::select(depmap_id, lineage) %>%
              dplyr::full_join(copyNumber, by = "depmap_id") %>%
              dplyr::filter(gene_name == "RPL14") %>%
              ggplot(aes(x = lineage, y = log_copy_number, fill = lineage)) +
        geom_boxplot(outlier.alpha = 0.1) +
        ggtitle("Boxplot of log genomic copy number values for gene RPL14 by lineage") +
        theme(axis.text.x = element_text(angle = 45, hjust=1)) +
        theme(legend.position = "none")



```