---
title: "Ex-6"
author: "Nurit Edri"
date: "12/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
expFile=system.file("extdata",
                    "leukemiaExpressionSubset.rds",
                    package="compGenomRData")

mat=readRDS(expFile)
```

```{r}
boxplot(mat)

scaledMat <- scale(mat)
boxplot(scaledMat)

logMat <- log2(mat)
boxplot(logMat)

scaledLogMat <- scale(logMat)
boxplot(scaledLogMat)
```

```{r}
library(pheatmap)

# set the leukemia type annotation for each sample
annotation_col = data.frame(
                    LeukemiaType =substr(colnames(mat),1,3))
rownames(annotation_col)=colnames(mat)


pheatmap(mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale = "none",
         clustering_method="ward.D",
         main = "mat without scaling")

pheatmap(scaledMat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale = "none",
         clustering_method="ward.D",
        main = "Scaled mat")

 pheatmap(scaledLogMat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale = "none",
         clustering_method="ward.D",
         main = "Log2 Scaled mat")
```
3. For the transformed and untransformed data sets used in the exercise
above, use the silhouette for deciding number of clusters using hierarchical
clustering.
```{r}
library(cluster)
set.seed(101)
pamclu=cluster::pam(t(mat),k=5)
plot(silhouette(pamclu),main=NULL)

pamclu1=cluster::pam(t(scaledMat),k=5)
plot(silhouette(pamclu1),main=NULL)

pamclu2=cluster::pam(t(scaledLogMat),k=5)
plot(silhouette(pamclu2),main=NULL)
```

### Dimension reduction

```{r}
par(mfrow=c(1,2))

# create the subset of the data with two genes only
# notice that we transpose the matrix so samples are 
# on the columns
sub.mat=t(mat[rownames(mat) %in% c("ENSG00000100504","ENSG00000105383"),])

# ploting our genes of interest as scatter plots
plot(scale(mat[rownames(mat)=="ENSG00000100504",]),
     scale(mat[rownames(mat)=="ENSG00000105383",]),
     pch=19,
     ylab="CD33 (ENSG00000105383)",
     xlab="PYGL (ENSG00000100504)",
     col=as.factor(annotation_col$LeukemiaType),
     xlim=c(-2,2),ylim=c(-2,2))

# create the legend for the Leukemia types
legend("bottomright",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

# calculate the PCA only for our genes and all the samples
pr=princomp(scale(sub.mat))


# plot the direction of eigenvectors
# pr$loadings returned by princomp has the eigenvectors
arrows(x0=0, y0=0, x1 = pr$loadings[1,1], 
         y1 = pr$loadings[2,1],col="pink",lwd=3)
arrows(x0=0, y0=0, x1 = pr$loadings[1,2], 
         y1 = pr$loadings[2,2],col="gray",lwd=3)


# plot the samples in the new coordinate system
plot(-pr$scores,pch=19,
     col=as.factor(annotation_col$LeukemiaType),
     ylim=c(-2,2),xlim=c(-4,4))

# plot the new coordinate basis vectors
arrows(x0=0, y0=0, x1 =-2, 
         y1 = 0,col="pink",lwd=3)
arrows(x0=0, y0=0, x1 = 0, 
         y1 = -1,col="gray",lwd=3)

# The arrows represent eigenvectors showing the direction of greatest variation.
```

1. Do PCA on the expression matrix using the princomp() function and then use the screeplot() function to visualize the explained variation by eigenvectors. How many top components explain 95% of the variation? [Difficulty:Beginner]

```{r}

pr=princomp(scale(mat))
screeplot(pr)

#PCA
library(stats)
library(ggplot2) 
# install.packages("ggfortify") ggfortify is needed to let ggplot2 know about PCA data structure.
library(ggfortify)

#compute PCA
pcaResults <- prcomp(t(logMat))
autoplot(pcaResults, data = annotation_col, colour = 'LeukemiaType')
```

2. In this exercise we use the Rtsne() function on the leukemia expression
data set. Try to increase and decrease perplexity t-sne, and describe the
observed changes in 2D plots. [Difficulty: Beginner]

```{r}
library("Rtsne")

set.seed(42) 
tsne_out <- Rtsne(t(mat),perplexity = 10) # Run TSNE
 #image(t(as.matrix(dist(tsne_out$Y))))
# Show the objects in the 2D tsne representation
plot(tsne_out$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19)

# create the legend for the Leukemia types
legend("bottomleft",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)

```
```{r}
set.seed(42) 
tsne_out <- Rtsne(t(mat),perplexity = 5) 
plot(tsne_out$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19)


```
```{r}

set.seed(42) 
tsne_out <- Rtsne(t(mat),perplexity = 15) 
plot(tsne_out$Y,col=as.factor(annotation_col$LeukemiaType),
     pch=19)

```

